Misexpression of germline programs with loss of the X-linked intellectual disability factor KDM5C

### 2024-05-30 submission/manuscript.Rmd
update text

```{r, DESeq2_params, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
#get the parameters used repeatedly, like the log2fold change cut off
source("../code/utilities/parameters.R")

```
RNA sequencing (RNA-seq) studies surprisingly revealed the aberrant expression of testis genes within the adult _Kdm5c_ knockout (-KO) hippocampus[@scandagliaLossKdm5cCauses2017]. Given the high abundance of testis-enriched genes within the mouse transcriptome, it is currently unclear if their misexpression in the _Kdm5c_-KO brain is indicative of random de-repression throughout the genome or instead reflects an impairment in tissue identity. To distinguish these possibilities, we globally assessed the expression of previously characterized tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017] in our previously published mRNA-seq datasets of the _Kdm5c_-KO amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020]

@@

We then aimed to globally characterize germline gene misexpression in _Kdm5c_-KO mice, but lacked a comprehensive list of mouse germline-enriched genes. To facilitate downstream analyses, we generated a list of male and female germline-enriched genes by evaluating gene expression in wild-type and germline-depleted (Kit^W/Wv^) mice. We assessed gene expression in male and female mice at embryonic day 12, 14, and 16 embryos, as well as male postnatal day 6 and adult testes. We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type germline 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of maximum expression in wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes.

>>>>

```{r, DESeq2_params, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
#get the parameters used repeatedly, like the log2fold change cut off
source("../code/utilities/parameters.R")
```

RNA sequencing (RNA-seq) studies surprisingly revealed the aberrant expression of testis genes within the adult _Kdm5c_ knockout (-KO) hippocampus[@scandagliaLossKdm5cCauses2017]. Given the high abundance of testis-enriched genes within the mouse transcriptome, it is currently unclear if their misexpression in the _Kdm5c_-KO brain is indicative of random de-repression throughout the genome or instead reflects an impairment in tissue identity. To distinguish these possibilities, we globally assessed the expression of previously characterized tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017] in our previously published mRNA-seq datasets of the _Kdm5c_-KO amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020]

@@

We then aimed to globally characterize germline gene misexpression in _Kdm5c_-KO mice, but lacked a comprehensive list of mouse germline-enriched genes. To facilitate downstream analyses, we generated a list of male and female germline-enriched genes by evaluating gene expression in wild-type and germline-depleted (Kit^W/Wv^) mice. We assessed gene expression in male and female mice at embryonic day 12, 14, and 16, as well as male postnatal day 6 and adult testes. We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type germline 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of maximum expression in wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes.
